Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-13 treestats_paper.tex
removed tsinfer

\newcommand{\tskit}{{\texttt{tskit}}}
\newcommand{\tsinfer}{{\texttt{tsinfer}}}
\newcommand{\branch}{\mbox{Branch}} % branch stat
\newcommand{\branchp}{\mbox{Branch}_+} % polarised
\newcommand{\site}{\mbox{Site}} % site stat
\newcommand{\sitep}{\mbox{Site}_+} % polarised
\newcommand{\node}{\mbox{Node}} % node stat
\newcommand{\nodep}{\mbox{Node}_+} % polarised
\newcommand{\given}{\;\vert\;}

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at scale.
In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct, which no inference method can hope to achieve.
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
in 1Mb windows separately in each of the five continental groupings,
calculated using the site statistic described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (of Example~\ref{ex:branch_diversity})
as calculated from three inferred tree sequences.
Firstly, we used the tree sequence estimated by \tsinfer{}
\citep{kelleher2019inferring}. By default, \tsinfer{} uses the frequency of the
derived allele at each site as a proxy for its age (and, therefore, the age
corresponding ancestor and node), aiming only to estimate accurate
tree topologies. We therefore expect the correspondence between branch
length and site statistics in this case to be poor. Secondly,
we used allele age estimates for each site supplied by GEVA~\citep{albers2019dating},
as input to \tsinfer{}. In this case, ancestor ages were determined by
the GEVA allele age estimates rather than the default frequency
based heuristic used by \tsinfer{}; inference was run as usual
otherwise. Thirdly, we used trees and node times estimated
by Relate~\citep{speidel2019method}.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
Separate curves for each of the 26 sampling locations are shown in Supplemental Figure \ref{fig:chr20_full}.

The comparisons are shown in Figure~\ref{fig:chr20}.
As expected, branch diversity from \tsinfer{} shows the worst agreement with
site diversity.
The two methods that infer node times differ substantially,
with Relate showing a significantly tighter agreement than
GEVA+\tsinfer{}.
Does this mean that Relate is doing a better job at inferring the tree sequence
than GEVA+\tsinfer?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

\jk{I'm not particularly happy with this to be honest: it's all very
hand-wavey. Can we be more precise about the correspondence between
site and branch statistics? Relate looks a bit better, but I don't
think we can say "significantly" (sounds like p-values or something).
Is there some sort of time-series analysis we could do to show how
similar sets of squiggly lines are to each other?}

\krt{
Overall, I have to squint at Fig S2 to see differences, and most of the issue seems to be near what I assume to be a
centromere?}

\begin{figure}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.region.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.region.diversity}.pdf}
    \caption{
        \textbf{(top left)} Mean sequence diversity
        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
        and the dual branch statistic as calculated from tree sequences inferred by
        \textbf{(top right)} \tsinfer,
        \textbf{(bottom left)} Relate, and
        \textbf{(bottom right)} GEVA+\tsinfer.
        Separate curves are plotted for each sampling location.
        \label{fig:chr20}
    }
\end{figure}

In practice, we do not expect branch and site diversity to agree exactly,
because of local differences in mutation rate and intensities of selected mutations.
For instance, regions in which a high proportion of potential mutations are deleterious
are expected to have a lower diversity for two reasons:
first, the deleterious mutations themselves are less likely to be found,
and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
and thus diversity at even neutral, linked sites \citep{hudson1994levels,charlesworth1997effects}.
The first effect would cause site and branch statistics to differ,
because it effectively reduces the mutation rate in the region
and violates the assumption of independence of mutations given the trees.
However, the second effect does \emph{not} affect the correspondence between site and branch statistics,
since it is mediated by tree shape.
However, it is generally unknown how much mutation rate varies along the genome
\citep[but see][]{supek2015differential}
or how dense are targets of selection \citep{leffler2012revisiting}.
For this reason, it is very interesting to see how close a correspondence between site and branch
statistics it is \emph{possible} to obtain --
it is tempting to say that the best tree sequence would obtain
as tight a match between site and branch statistics as possible,
with remaining variation explained by direct selection and/or mutation rate variation.

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Time stratification of statistics}

Thus far we have examined how statistics vary along the genome,
but a tree sequence allows us access to another dimension: time.
Next, we examine the \emph{time} distribution of admixture signal
in the tree sequence inferred by GEVA+\tsinfer.
The Branch $f_4(A, B; C, D)$ statistic for four sample sets $A$, $B$, $C$, and $D$
is the sum of $(p_A - p_B)(p_C - p_D)$ across branches, multiplied by span and length,
where $p_X$ is the proportion of samples in sample set $X$ that inherit from that node.
A branch with a large value is expected to contribute a large amount to the Site $f_4$,
which is calculated from genotyping data
and is used to infer historical connections between populations.
To visualize the time signature, we assigned the contribution of each branch
to its terminal node,
and then summed these values across all nodes in 50 logarithmically-spaced bins by time.
We then normalised these to sum to 1 by dividing by the total Branch $f_4$ value.

The four-tuple of sample sets for the first two statistics (PUR,TSI;GWD,JPT and ASW,CEU;MSL,CHB)
are from similar regions of the world,
and show almost identical time signatures,
even though the second set of sample sets has a value of $f_4$ nearly eight times higher.
The second two statistics (IBS,GBR;FIN,JPT and TSI,CEU;FIN,CHB) are also comparable,
and also show similar signals to each other.
These curves are noisier, because the $f_4$ values are much smaller.
For all four comparisons, the largest contribution to $f_4$ comes from
ancestors living tens of thousands of generations ago.
\plr{WHAT IS THE TIME UNIT?}
This is surprising, because $f_4$ is expected to deviate from zero due to asymmetries in tree topology
during recent history (e.g., more branches ancestral to ASW and MSL than CEU and MSL).
Furthermore, we expected the different sample sets to have much more distinct time signatures.
The shared, ancient time signatures may be because nonzero $f_4$ statistics
are due to differentially sorting ancestral variation,
or inacurracies in tree sequence inference.
It is tempting but propably unwarranted to attribute the
large downward spike between 2000 and 3000 generations in the TSI,CEU;FIN,CHB comparison
to a historical connection between the ancestors of CEU and the Finnish not shared by the Tuscans
during that time period.
Making rigorous conclusions from this sort of visualization
will require substantial validation work
to understand the various sources of bias and uncertainty.

\begin{figure}
    \includegraphics{{figures/1kg/tgp_geva_chr20.node.time.f4}.pdf}
    \caption{
        Time distribution of signal for four $f_4$ statistics
        in the tree sequence for chromosome 20 inferred from 1000 genomes data
        by GEVA+\tsinfer{}.
        The branch $f_4$ statistics for these four are:
        PUR,TSI;GWD,JPT: 1603,
        ASW,CEU;MSL,CHB: 12587,
        IBS,GBR;FIN,JPT: -129, and
        TSI,CEU;FIN,CHB: -96.
        Population codes are
        `PUR': Puerto Ricans from Puerto Rico;
        `TSI': Toscani in Italia;
        `GWD': Gambian in Western Divisions in the Gambia;
        `JPT': Japanese in Tokyo, Japan;
        `ASW': Americans of African Ancestry in SW USA;
        `CEU': Utah Residents (CEPH) with Northern and Western European Ancestry,
        `MSL': Mende in Sierra Leone;
        `CHB': Han Chinese in Beijing, China;
        `IBS': Iberian Population in Spain;
        `GBR': British in England and Scotland;
        and `FIN': Finnish in Finland.
        \label{fig:node_f4}
    }
\end{figure}

@@

@@

The duality between Site and Branch statistics depends crucially on two assumptions:
that each new mutation is distinguishable, as in the infinite-sites model,
and that mutation rates are constant along the genome.
Neither of these things are true, but are they good approximations?
Multiple mutations at the same site, causing back mutations,
are present in real datasets but are generally thought to be uncommon,
especially if hypervariable sites are excluded.
The effect of multiple mutations on Site statistics
can be bounded by a function of the proportion of sites at which these occur.
Large-scale variation in mutation rate (at, say, the megabase scale)
might plausibly have a stronger effect on the correspondence,
because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
then dividing the Site statistic by the Branch statistic
would give us an estimate of relative average mutation rate in windows along the genome.
Similarly, Branch statistics from well-calibrated tree sequences
would provide a way to assess variation in population genetic summaries along the genome
in a way independent of local mutation rate.

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.site.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/1kg_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/tgp_geva_chr20.branch.1000000.diversity}.pdf}
    \includegraphics[width=0.5\textwidth]{{figures/1kg/relate_chr20.branch.1000000.diversity}.pdf}
    \caption{
        \textbf{(top left)} Mean sequence diversity
        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
        and the dual branch statistic as calculated from tree sequences inferred by
        \textbf{(top right)} \tsinfer,
        \textbf{(bottom left)} GEVA, and
        \krt{This should be tsinfer+GEVA, no?}
        \textbf{(bottom right)} Relate.
        \label{fig:chr20_full}
    }
\end{figure}

@@

%%%%%%%%%%%%%%%%%%
\subsection*{Genomic PCA}

Let $G$ denote the genotype matrix, so that $G_{ja} \in \{0,1\}$
is the genotype of the $a^\text{th}$ individual
at the $j^\text{th}$ site, with `1` denoting the derived state
(we assume here biallelic markers).
Also let $P_j = (1/n) \sum_a G{ja}$ denote the vector of allele frequencies.
Then the \emph{genetic covariance} between samples $a$ and $b$ is
\begin{align*}
    C_{ab}
        &= \frac{1}{L} \sum_j (G_{ja} - P_j) (G_{jb} - P_j) . %  \\
        % &= \frac{1}{L} \left( (G - P \bone^T)^T (G - P \bone^T) \right)_{ab} .
\end{align*}
This can be computed by letting
$\iw_a = \delta_a$ and $\iw_b = \delta_b$
(these weights mark nodes ancestral to $a$ and $b$ respectively)
and $\iw_t = \bone / n$
(this weight gives the proportion of samples inheriting from each node).
Then, with
$f(\nw_a, \nw_b, \nw_t) = (\nw_a - \nw_t) (\nw_b - \nw_t)$,
we can compute the covariance as
\begin{align*}
    C_{ab} = \frac{1}{2} \bar \site(f, (\iw_a, \iw_b, \iw_t)) .
\end{align*}
This follows, including the factor of two,
because the summary function applied to the ancestral allele
at a site with derived allele frequencies $p_a$, $p_b$, and $p_t$
in sample $a$, sample $b$, and total, respectively, is
$f(1 - p_a, 1 - p_b, 1 - p_t) = f(p_a, p_b, p_t)$.

Computing the full covariance matrix for a large number of samples may be infeasible.
However, we can efficiently find $y^T C z$ for arbitrary vectors $y$ and $z$.
This works because if we set the initial weights to $y$,
then each allele's weight is equal to the sum of the entries of $y$
corresponding to samples carrying that allele.
Then, if we define
$f_{yz}(u, v, t) = (u - (\sum_a y_a) t) (v - (\sum_b z_b) t)$,
then
\begin{align} \label{eqn:yCz}
    \frac{1}{2} \bar \site(f_{yz}, (y, z, \iw_t))
        &= \frac{1}{L} \sum_j (\sum_a y_a G_{ja} - (\sum_a y_a) P_j)
                    (\sum_b z_b G_{jb} - (\sum_b z_b) P_j) \\
        &= y^T C z .
\end{align}

\plr{Hmph: to actually compute $Cz$ we need one weight vector for each sample, again.
    TODO: look up if you can do Krylov just by computing $y^t C z$ for a reasonable number of vectors.
    We could compute $Cz$ quickly if we could propagate down the tree:
    we'd have $(Cz)_a$ equal to the sum of the weights at all nodes \emph{ancestral to} $a$. }

Where does the signal from an eigenvector come from?
Let $y$ be an eigenvector of $C$ with eigenvalue $\lambda$, normalized so that $\|y\|^2 = 1$.
Since $Cy = \lambda y$,
we can write $1 = \sum_a y_a^2 = (1/\lambda) y^T C y$.
However, equation \eqref{eqn:yCz} gives $y^T C y$ as a sum over \emph{sites},
and thus distributes the variance in $G$ associated with $y$
across the tree sequence.
(This is true because of the singular value decomposition of $G$.)

>>>>

\newcommand{\tskit}{{\texttt{tskit}}}
\newcommand{\branch}{\mbox{Branch}} % branch stat
\newcommand{\branchp}{\mbox{Branch}_+} % polarised
\newcommand{\site}{\mbox{Site}} % site stat
\newcommand{\sitep}{\mbox{Site}_+} % polarised
\newcommand{\node}{\mbox{Node}} % node stat
\newcommand{\nodep}{\mbox{Node}_+} % polarised
\newcommand{\given}{\;\vert\;}

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at scale.
In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct, which no inference method can hope to achieve.
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated mean sequence diversity along chromosome 20 from the 1000 Genomes dataset
in 1Mb windows separately in each of the five continental groupings,
calculated using the site statistic described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (of Example~\ref{ex:branch_diversity})
from the trees and node times estimated by Relate~\citep{speidel2019method}.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
Does this mean that Relate is doing a better job at inferring the tree sequence
Is this an indication that Relate is doing a good job at inferring trees,
or just a byproduct of the inference method (which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

\begin{figure}
    \includegraphics[width=0.8\textwidth]{{figures/1kg/relate_chr20.site.1000000.region.diversity}.pdf}
    \includegraphics[width=0.8\textwidth]{{figures/1kg/relate_chr20.branch.1000000.region.diversity}.pdf}
    \caption{
        \textbf{(top)} Mean sequence diversity
        in 1Mb windows along human chromosome 20 of the 1000 Genomes Project,
        and \textbf{(bottom)}
        the dual branch statistic as calculated from tree sequences inferred by Relate~\citep{speidel2019method}.
        Separate curves are plotted for each sampling location.
        \label{fig:chr20}
    }
\end{figure}

In practice, we do not expect branch and site diversity to agree exactly,
because of local differences in mutation rate and intensities of selected mutations.
For instance, regions in which a high proportion of potential mutations are deleterious
are expected to have a lower diversity for two reasons:
first, the deleterious mutations themselves are less likely to be found,
and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
and thus diversity at even neutral, linked sites \citep{hudson1994levels,charlesworth1997effects}.
The first effect would cause site and branch statistics to differ,
because it effectively reduces the mutation rate in the region
and violates the assumption of independence of mutations given the trees.
However, the second effect does \emph{not} affect the correspondence between site and branch statistics,
since it is mediated by tree shape.
It is generally unknown how much mutation rate varies along the genome
\citep[but see][]{supek2015differential}
or how dense are targets of selection \citep{leffler2012revisiting}.
For this reason, it is very interesting to see how close a correspondence between site and branch
statistics it is \emph{possible} to obtain --
it is tempting to say that the best tree sequence would obtain
as tight a match between site and branch statistics as possible,
with remaining variation explained by direct selection and/or mutation rate variation.

@@

Many genomic datasets only contain genotype information at a subset of sites
that do not represent an unbiased look at genome-wide variation,
introducing an ascertainment bias that difficulty for many applications.
If, however, unbiased tree sequences can be inferred using these data,
then Branch statistics could provide unbiased estimates of genome-wide quantities.
This is in principle possible, as \citet{kelleher2019inferring} inferred tree sequences
with UK BioBank data, produced using genotyping arrays \citep{bycroft2018genome}.
This would be similar to imputation, and would likely face similar challenges.

@@

The duality between Site and Branch statistics depends crucially on two assumptions:
that each new mutation is distinguishable, as in the infinite-sites model,
and that mutation rates are constant along the genome.
Neither of these things are true, but are they good approximations?
Multiple mutations at the same site, causing back mutations,
are present in real datasets but are generally thought to be uncommon,
especially if hypervariable sites are excluded.
The effect of multiple mutations on Site statistics
can be bounded by a function of the proportion of sites at which these occur.
Large-scale variation in mutation rate (at, say, the megabase scale)
might plausibly have a stronger effect on the correspondence,
because a Branch and Site statistic are equal in expectation up to multiplication by the mutation rate.
For instance, if we knew that the inferred tree sequences of \autoref{fig:chr20} were correct,
then dividing the Site statistic by the Branch statistic
could give us an estimate of relative average mutation rate in windows along the genome.
Similarly, Branch statistics from well-calibrated tree sequences
would provide a way to assess variation in population genetic summaries along the genome
in a way independent of local mutation rate.

@@

@@
